Precise temporal localisation of M/EEG effects with Bayesian generalised additive multilevel models

### 2025-08-27 manuscript.qmd
pushing first public version of the manuscript

In brief, our results show that the model-based approach we introduced outperforms conventional cluster-based methods in identifying the onset and offset of M/EEG effects. We first assessed the performance of this approach on simulated data allowing us to evaluate the method's ability to recover ground-truth onset and offset values. We then assessed its performance on actual MEG data, allowing us to assess its sensitivity to realistic data properties (subset similarity). Together, these results highlight desirable properties for any method aiming to precisely and reliably estimate the onset and offset of M/EEG effects: it should i) recover true onsets and offsets in simulation (good asymptotic behaviour), ii) identify clusters that are interpretable and consistent in empirical data, and iii) show sensitivity to subtle changes in the data. Our approach meets all three of these desiderata.

@@

<!--

We chose not to include the cluster-depth algorithm [@frossard2022], as prior work already shows its performance to be inferior to cluster-mass approaches in similar contexts [@rousselet_using_2025].

To make this paragraph more concrete, can we reflect on our own method in this regard?

-->

@@

<!--

$$
\begin{aligned}
\text{AUC}_{i} &\sim \mathrm{Beta}(\mu_{i}, \phi)\\
g(\mu_{i}) &= f \left(\text{train}_{i}, \text{test}_{i} \right)\\
\end{aligned}
$$

where we assume that AUC values come from a $\mathrm{Beta}$ distribution with two parameters $\mu$ and $\phi$. We can think of $f \left(\text{train}_{i}, \text{test}_{i} \right)$ as a surface (a smooth function of two variables) that we can model using a 2-dimensional splines. Let $\mathbf{s}_{i} = \left(\text{train}_{i}, \text{test}_{i} \right)$ be some pair of training and testing samples, and let $\mathbf{k}_{m} = \left(\text{train}_{m}, \text{test}_{m} \right)$ denote the $m^{\text{th}}$ knot in the domain of $\text{train}_{i}$ and $\text{test}_{i}$. We can then express the smooth function as:

$$
f \left(\text{train}_{i}, \text{test}_{i} \right) = \alpha + \sum_{m=1}^M \beta_{m} b_{m} \left(\tilde{s}_{i}, \tilde{k}_{m} \right)
$$

A popular bivariate basis function uses *thin-plate splines* [@wood2003]. These splines are designed to interpolate and approximate smooth surfaces over two dimensions (hence the "bivariate" term). For $d=2$ dimensions and $l=2$ (smoothness penalty involving second order derivative):

$$
f \left(\tilde{s}_{i} \right) = \alpha + \beta_{1} x_{i} + \beta_{2} z_{i} +\sum_{m=1}^{M} \beta_{2+m} b_m\left(\tilde{s}_i, \tilde{k}_m\right)
$$

using the the radial basis function given by:

$$
b_m\left(\tilde{s}_i, \tilde{k}_m\right)=\left\|\tilde{s}_i-\tilde{k}_m\right\|^2 \log \left\|\tilde{s}_i-\tilde{k}_m\right\|
$$

where $\left\|\mathbf{s}_i-\mathbf{k}_{m}\right\|$ is the Euclidean distance between the covariate $\mathbf{s}_{i}$ and the knot location $\mathbf{k}_{m}$.

$$
\begin{aligned}
\text{AUC}_{i} &\sim \mathrm{Beta} \left(\mu_{i}, \phi\right)\\
\operatorname{logit} \left(\mu_{i} \right) &= \alpha + \sum_{k=1}^K b_{k} B_{k}\left(\text{train}_{i}, \text{test}_{i} \right)
\end{aligned}
$$

-->

>>>>

In brief, our results show that the model-based approach we introduced outperforms conventional cluster-based methods in identifying the onset and offset of M/EEG effects. We first assessed the performance of this approach on simulated data, allowing us to evaluate the method's ability to recover ground-truth onset and offset values. We then assessed its performance on actual MEG data, allowing us to assess its sensitivity to realistic data properties (subset similarity). Together, these results highlight desirable properties for any method aiming to precisely and reliably estimate the onset and offset of M/EEG effects: it should i) recover true onsets and offsets in simulation (good asymptotic behaviour), ii) identify clusters that are interpretable and consistent in empirical data, and iii) show sensitivity to subtle changes in the data. Our approach meets all three of these desiderata.

@@

@@
